\documentclass[a4paper]{article}

\usepackage[left=1cm,top=2cm,right=1cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{natbib}
\bibliographystyle{apalike}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{placeins}
\pagestyle{fancy}
\lhead{\today}

\newcommand{\koc}{$\mbox{K}_{\mbox{\textsc{oc}}}$}
\newcommand{\kclay}{$\mbox{K}_{\mbox{clay}}$}
\newcommand{\kd}{$\mbox{K}_{\mbox{d}}$}

%% \includeonly{}

\begin{document}

\section*{Daisy simulation of Silstrup and Estrup}

The Danish Pesticide Leaching Assessment Programme (PLAP) has been
monitoring drain and soil water since 1999 in six different locations
in order to evaluate the leaching risk of pesticides.  The drain water
measurements pesticides are found in some drain water measurements,
but not in the horizontal filters 3.5 meter below surface.

In order to better understand the system, and eventually how the
measurements can be better used for assessing potential risk of
contamination of drinking water, the Daisy agricultural model has been
extended with support for those processes we assume are relevant for
transport of pesticides from surface to drain pipes.

To test our understanding as embedded in the model, as well as the
applicability for the PLAP sites, a pilot project with two PLAP sites,
four pesticides, and two years of data have been modelled.  Our
hypothesis is here that we can explain the measured data with the
model.

The most significant measured results are from the Estrup site, so
that was fixed as the first site.  We wanted to use the same years for
both sites to see the results of similar climate on two different
locations.  Hourly weather data from the site had to be present, both
for the modelling period and one growth season before for warmup.  The
initial choice of F{\aa}rdrup for the second site was rejected as it
was not possible to get sufficient quality weather data.  Furthermore,
we wanted the same pesticides on both sites, and both weakly and
strongly sorbing pesticides represented.  Glyphosate had to be one of
them.

Based on these criteria, we chose Silstrup and Estrup, drain seasons
2000-2001 and 2001-2002, with the pesticides glyphosate, fenpropimorph,
dimethoate, and metamitron.

Both sites are described in detail in \citet{lindhardt2001}, with an
overview of the measured data found in \citet{vap2009}.  Estrup is a
pedologically rich site, containing both areas with sand and clay,
peat, thin layers of chalk, and even sand from railroad construction.
Silstrup is also heterogeneous, but less so, with high levels of clay
dominating the area.

\section{Model setup}

\subsection{Weather}

Hourly weather data for Silstrup, Tylstrup and Askov (near Estrup) was
provided by the Faculty of Agricultural Sciences, Aarhus University.
The idea was that Tylstrup data could be used to supplement holes in
the Silstrup data.  Both the Silstrup and Askov data sets contained
several short holes.  We filled those by using values from the
preceding or following hours.  The Silstrup data set ended at
2002-03-12.  The drain season ended 2002-03-20, with 0.6 mm water
collected from the drains the last 8 days.  Baring this in mind, we
chosed to end the simulation 2002-03-11, rather than continue with
data from another station.

The weather data contained air temperature, wind speed, relative
humidity, precipitation, and global radiation are used.  Based on
these data, Daisy can use FAO version of Penman-Monteith equation
\citep{FAO-PM} to calculate reference evapotranspiration (ET0).  From
that, Daisy will calculate potential evapotranspiration (ETc) by using
the crop leaf area index (LAI) with Beer's law to dividing the surface
into canopy and bare soil, and using different factors for each
fraction.  Based partly on \citet{kjaersgaard2008crop}, crop factor of
1.2 and a bare soil factor of 0.6 was chosen.  Actual
evapotranspiration (ETa) is further limited by the capability of the
root system and soil surface to extract water from the soil.

Precipitation, air temperature, ETc and ETa can be seen on
figures~\ref{fig:Silstrup-weather} and~\ref{fig:Estrup-weather}.

\subsection{Management}

All management data were provided by GEUS.

\subsubsection{Tillage}

Date, type, and depth were specified for all tillage operations.  All
three were entered into Daisy.  In Daisy, the main effects of tillage
are to incroporate some of the surface material into the soil
(depending on the type of tillage operation), and to mix the content
of the soil in the specified depth.  This is not expected to affect
pesticide leaching much.  In the real world, the main effect of
tillage applicable to pesticide leaching is likely to be a change in
the hydraulic properties for the top soil resulting from the tillage
operation.  Since we chose to implement dynamic hydraulic soil
properties for the Silstrup soil surface, the tillage information were
useful there as well.

\subsubsection{Fertilization}

Date, type, and amount were specified for each fertilization event, as
well as N, P and K content of fertilizer.  Of these nutrients, Daisy
can normally only handle N, and N has been disabled for these
simulations to save time.  All fertilization event have been added to
Daisy, but without N there will be no effect of the mineral
fertilizer.  The organic fertilizer will have a minimal effect, the
water content of the fertilizer will be added, and the dry matter
content will be added to the litter where it can catch water and
pesticides, until it become incroporated into the soil by either
tillage operations or earthworm activity.  However, due to the timing
of the applications, both effect is likely to be negigible.

\subsubsection{Crop management}

For sowing information about date, crop and sowing density were
provided.  The Daisy crop parametrizations do not rely on sowing
density, but instead assume that ``standard practise'' is used.  Daisy
has experimental support for a crop model that include sowing density,
but given that the PLAP sites are expected to follow standard
management practise, we found it safer to use the older and better
tested crop parameterizations.

Information about the crow growth was given phenologically (BBCH
stage) and in terms of above ground biomass. No attempt was made
calibrate the crop in order to match this with the simulated
development stage and biomass.  Information about the two important
parameters for the water balance, namely leaf area and root density,
were not provided, but are normally related to the two measured
parameters.

Harvest data included date, stubble height, as well as grain and straw
yields.  Date and stubble height can be directly used by Daisy.  Grain
yield can be used for calibration, however, as crop production is not
the focus of this project, we merely noted that both measured and
simulated yield where within the normal range.  The ratio of between
grain and straw yield was used for a coarse estimation of the fraction
of the crop left on the field as residuals after harvest.  The
residuals play a crusial role in the simulation for the Silstrup
glyphosate leaching.

\begin{table}[htbp]
  \caption{Crop management.}
  \label{tab:crop-man}
  \centering
  \begin{tabular}{c||c|c|c||c|c|c}\hline
    & \multicolumn{3}{c||}{Silstrup} & \multicolumn{3}{c}{Estrup}\\
    Year & Crop          & Sow   & Harvest & Crop          & Sow   & Harvest\\
\hline
    2000 & Fodder Beet   & 04/05 & 15/11   & Spring Barley & 12/4  & 28/8 \\
    2001 & Spring Barley & 9/5   & 5/9     & Peas          & 5/2   & 22/8 \\
    2001 &               &       &         & Winter Wheat  & 19/10 & --- \\
  \end{tabular}
\end{table}

The only calibration done on crop management was to replace the fodder
beet (see table~\ref{tab:crop-man}) with spring barley.  The Silstrup
soil water measurements indicated that we underestimated ability of
the summer 2000 crop to empty the plow layer soil water (see
figure~\ref{fig:Silstrup-theta}).  The spring barley parametrization
did a better job at this than the less tested fodder beet.

\subsubsection{Pesticide and bromide application}

The data for pesticide application consisted of date, amount, and
trade name.  Trade name is is translated to active ingredient using
``Middeldatabasen'' from \textsc{dlbr} Landbrugsinfo
(\url{http://www.landbrugsinfo.dk/}).  Note that Goltix WG application
is given in L/ha, while the active ingredient is specified in g/kg.
To resolve, a density of 1~kg/L is assumed.  For potassium bromide,
bromide content was calculated from relative molar mass.

The applications are sumarized in table~\ref{tab:man-pest}.
\begin{table}[htbp]
  \caption{Pesticide and bromide application.  Only those active 
    ingredients we track are listed.  }
  \label{tab:man-pest}
  \centering
  \begin{tabular}{c|crl|crl}\hline
    \emph{Silstrup} & Trade name & \multicolumn{2}{c|}{Amount} 
               & Active ingredient & \multicolumn{2}{c}{Active amount} \\
    \hline
    2000-05-22 & Potassium bromide & 30 & kg/ha & Bromide    & 20.14 & kg/ha\\
    2000-05-22 & Goltix WG      & 1   & L/ha & Metamitron    & 700   & g/ha\\
    2000-06-15 & Goltix WG      & 1   & L/ha & Metamitron    & 700   & g/ha\\
    2000-07-12 & Goltix WG      & 1   & L/ha & Metamitron    & 700   & g/ha\\
    2001-06-21 & Tilt Top       & 0.5 & L/ha & Fenpropimorph & 187.5 & g/ha\\
    2001-07-04 & Tilt Top       & 0.5 & L/ha & Fenpropimorph & 187.5 & g/ha\\
    2001-07-16 & Perfektion 500 & 0.6 & L/ha & Dimethoate    & 300   & g/ha\\
    2001-10-25 & Roundup Bio    & 4.0 & L/ha & Glyphosate    & 1440  & g/ha\\
    \hline\hline
    \emph{Estrup} & Trade name & \multicolumn{2}{c|}{Amount} 
               & Active ingredient & \multicolumn{2}{c}{Active amount} \\
    \hline
    2000-05-15 & Potassium bromide & 30 & kg/ha & Bromide & 20.14 & kg/ha\\
    2000-06-15 & Tilt Top       & 0.5 & L/ha & Fenpropimorph & 187.5 & g/ha\\
    2000-06-15 & Perfektion 500 & 0.4 & L/ha & Dimethoate    & 200   & g/ha\\
    2000-07-05 & Tilt Top       & 0.5 & L/ha & Fenpropimorph & 187.5 & g/ha\\
    2000-06-15 & Perfektion 500 & 0.4 & L/ha & Dimethoate    & 200   & g/ha\\
    2000-10-13 & Roundup Bio    & 4.0 & L/ha & Glyphosate    & 1440  & g/ha\\
  \end{tabular}
\end{table}

\subsection{Pesticide and bromide properties}

Of the four pesticides examined, only metamitron and glyphosate were
memasured in concentrations above the detection limit in the examined
dataset.  A single sample at the detection limit were found for
dimethoate, and none were found for fenpropimorph.  No calibration has
therefore been performed on those.  

\subsubsection{Soil sorption and decomposition}

Sorption and decomposition parameters for pesticides are primarily
taken from \citet{ppdb20100517}, with values as shown in
table~\ref{tab:ppdb}.  The database specify a \koc{} value
independently of whether the pesticide is actually sorbed to organic
matter.  We chosed to use a \kd{} values measured in Denmark for the
two main pesticides.  For metamitron, \citet{madsen2000pesticide}
specify \kd{} values together with sil properties for a number of
Danish sites.  Section~\ref{sec:cal-metamitron} describes how we
selected a \kd{} based on those.  For glyphosate, the \kd{} value is
from \citet{gjettermann2009}.  The asorption is not instantaneous, an
adsorption rate of 0.05 h$^-1$ was used as an reasonable initial
guess.  We found no reason to change the value, except for glyphosate,
see section~\ref{sec:cal-glyphosate}.

The effect of depth on decomposition is taken from
\citet{focus2000,focus2002}.  The build in effect of temperature and
humidity for turnover of organic matter in Daisy is also applied for
pesticides.  The default diffusion coefficient used by Daisy for
pesticides of 4.6e-6 cm$^2$/s is used unchanged for all pesticides as
well as for colloids.  For bromide, a value of 2.0e-5 cm$^2$/s is used
instead, reflecting the smaller molecules.  We assume that the
pesticide molecules are all reflected by the roots, so there is no
crop uptake of pesticides.

\begin{table}[htbp]
  \centering
  \caption{Pesticide properties from  \citet{ppdb20100517}. 
    The DT50 value is the decomposition halftime in days.  For both DT50
    and \koc{} we put the value marked  'field' in \citet{ppdb20100517}
    in the center, surrounded by the lower and upper limit found in 
    field studies, as marked in a note in the database. For fenpropimorph
    the \koc{} field value did not fall within the specifed interval.  
    The \kd{} value for glyphosate is from \citet{gjettermann2009}, and
    the \kd{} range for metamitron is from \citet{madsen2000pesticide}.  
    The values used are in \textbf{bold}.}
  \label{tab:ppdb}

  \begin{tabular}{l|ccc}\hline
    Name & DT50 & \koc{} [ml/g] & \kd{} [ml/g] \\\hline
    Dimethoate & 4.6 -- \textbf{7.2} -- 9.8 & 16.25 -- \textbf{30} -- 51.88 & \\
    Fenpropimorph & 8.8 -- \textbf{25.5} -- 50.6 & 2771 -- \textbf{2401} -- 5943 &\\
    Glyphosate & 5 -- \textbf{12} -- 21 & 884 -- 21699 -- 60000  & \textbf{503} \\
    Metamitron & 6.6 -- \textbf{11.1} -- 22.0 & 77.1 -- 80.7 -- 132.5 & 0.14 -- \textbf{4.0}\\
  \end{tabular}
\end{table}

\subsubsection{Surface turnover}

None of our sources had specific information on above ground turnover.
As the glyphosate calibration (see section~\ref{sec:cal-glyphosate})
depend on part of the glyphosate staying in the litter pack for
several days, surface turnover potentially becomes a factor.  The
default value in Daisy of DT50 = 3.5 for pesticides is used.

\subsubsection{Colloids and colloid facilitated transport}

We have no data for colloids, so the parameters for colloid generation
and filtering calibrated for Rørrendegård have been reused for both
sites.  The model itself will adjust to the different clay content.
The pesticides are assumed to be able to sorb to and be transported
with colloids, meaning the colloids will be in competition with the
soil matrix as potential sorbtion sites for the solute form of the
pesticides.  In the lack of any measurements, we have assumed that the
colloids are 10000 more likely to attack to a dissolved soil particle
(a colloid) than to a stationary soil particle (part of the soil
matrix).

\subsubsection{Glyphosate calibration}
\label{sec:cal-glyphosate}

The highest glyphosate contrations in both sites were seen right after
application, or at the first large rain event after application.  This
is unlikely to a function of the pesticide properties, but rather of
the transport routes to the drain.  See~\ref{sec:cal-silstrup-surface}
for how this was calibrates.

The initial simulations showed practically no glyphosate movement once
it entered the soil matrix.  The measurements, however, did show some
late findings of glyphosate in the drain water.  In order to give the
glyphosate a chance to move, we divided the sorption into a weak but
fast and a strong but slow form.  The strong but slow form represent
90\% of the \kd{} value, the weak but fast form the remaining 10\%.
The effect is that the glyphosate is relatively mobile in the
beginning, but becomes less so as more glyphosate become sorbed in the
slow form.

\subsubsection{Metamitron calibration}
\label{sec:cal-metamitron}

Adjusting the decompose rate for metamitron has little effect on the
simulation results.  Figure~\ref{fig:Silstrup-C-Metamitron-2000} and
\ref{fig:Silstrup-Metamitron-2000-horizontal} shows why. The
metamitron we find in the drains is the same metamitron that was first
transported vertically to the end of the biopores, and then
horizontally towards the drain.  Since the biopores in the simulation
ends 1.2 meter below surface, this means the metamitron is located
below the 1 meter depth limit for decomposition specified by FOCUS.

In \citet{madsen2000pesticide} er sorbtionegenskaber bestemt for en
række danske lokaliteter.  Den bedste korrelation for sorption til
jordparametre er total jernoxid (FeO$_{\mbox{total}}$), og kun ringe
korrelation til organisk stof og ingen til let ekstraherbar jernoxid
(FeO$_{\mbox{oxalate}}$) som ellers er målt på Silstrup.  Den højeste
målte Kd 3,1 $\pm$ 0,9 L/kg for Drengsted, og den laveste 0,16 $\pm$
0,02 L/kg.  Vi regner derfor med at at Kd bør holde sig i intervalet
0,14 -- 4,0 L/kg.

En Kd på den højeste værdi i intervallet, 4,0 L/kg, giver det bedste
match for den tidlige hændelse.  Lavere værdier giver højere
udvaskning.  Vi kunne alternativt skrue ned på overfladeparametrene
for at lukke af for den hurtige transportvej ned, hvilket kunne
forsvares med at udbringningen er kort efter en jorbehandling, hvor vi
måske undervurderer pløjelagets ledningsevne, og overvurderer antallet
af bioporer.  Bromidmålingerne supporterer ikke dette.  Bromid blev
udbragt samtidig med metamitron, men der er desværre ikke nogen
målinger fra dræn i samme periode til at sammenligne med.  Men
eftersom der bliver observeret bromid i jordvand i en meters dybde må
vi konkludere at der findes nogen hurtige veje derned.  Faktisk tyder
målingerne i 1 og 3,5 meters dybde på at vi undervurderer hvor hurtigt
Bromid trænger ned, samtidig med at vi overvurderer hvor hurtigt
Metamitron trænger ned i dræn.

For at få den sene udvaskning af metamitron på plads kalibrerede vi os
frem til en halveringstid på 35 dage.

\subsubsection{Bromide calibration}
\label{sec:cal-bromide}

crop uptake reclection factor

\subsection{Soil}

\begin{verbatim}

** Lindhart, MACRO

** Biopores, cracks

** OM \& N

** Surface, plow pan

grundvand

\end{verbatim}

\subsubsection{Silstrup surface}
\label{sec:cal-silstrup-surface}

Glyphosate

Bromide

\subsection{Groundwater table}


\section{Results}

% \section*{Daisy simulation figures for Silstrup og Estrup}

\include{figs-Silstrup}

\include{figs-Silstrup-2D}

\include{figs-Estrup}

\include{figs-Estrup-2D}

\section{Discussion}

\begin{verbatim}
* Silstrup drain period may be too short due to heterogenity.

* Silstrup drain flow follow first rain, indicate surface flow and anisotropy

* Silstrup Metamitron not p{\aa}virket af halverinsgtid
\end{verbatim}

\section{Conclusion and further work}

It is possible to explain the measured data based on the processes
included in the present model, with some caveats
\begin{itemize}
\item The high degree of heterogeneity found in the Estrup site would
  require a detailed 3D model of the entire area to model
  mechanistically, the current 2D model setup can at best be viewed as
  ``effective parameters''.
\item The drain are activated more abruptly (first year) and sooner
  (second year) in reality, than what we have been able to simulate
  for the Silstrup site.  This is particularly noticeable for the
  Bromide measurements.  
\item Bromide is found in some of the horizontal filters at 3.5 meters
  depth at both sites in the first measurements after application of
  Bromide.  No pesticides are found in those depth though.  It does
  indicate a transport way for non-sorbing solutes that we cannot
  model.  One possibility is large scale fractures, this suggestion is
  supported by other work at GEUS.
\end{itemize}

Furthermore, the work presented in thus report cannot count as a
validation of either the parametrization or the conceptual model,
there are too many unknowns where we have had to guess or calibrate
parameters, and there are likely many different setups that would have
resulted in as good or better fits to the measured results.  It is not
certain that a proper validation is possible, but by applying the
model on more data sets we should be able to gain confidence in it it.

Apart from applying the model on more and larger data sets, two areas
in particular need further work.
\begin{itemize}
\item The surface processes (flow, litter storage, decomposition) are
  very important, especially for the Silstrup site, but only the
  minimal work on those to get an effect have been done in this project.
\item The pesticide processes (colloid transport, sorbtion sites,
  sorption kinetics, and decomposition) are parametrized from
  literature values and ``best guesses'', many of those parameters
  should be adapted to direct local measurements.
\end{itemize}

\addcontentsline{toc}{section}{\numberline{}References}
\bibliography{../../txt/daisy}

\end{document}
