\documentclass[a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage{csquotes}
\usepackage{hyperref}
\usepackage[sorting=ydnt,style=authoryear,dashed=false,maxnames=99,backend=biber]{biblatex}
\addbibresource{daisy.bib}

\input{macro}
\setcounter{secnumdepth}{-1} 

\newcommand{\daisykey}[2]{
\defbibfilter{daisy-#1}{keyword=daisy and keyword=#1}
\printbibliography[heading=subbibliography,%
filter=daisy-#1,title={#2}]
}
\newcommand{\daisytype}[2]{
\defbibfilter{daisy-#1}{keyword=daisy and type=#1 and not keyword=popular and not keyword=report}
\printbibliography[heading=subbibliography,%
filter=daisy-#1,title={#2}]
}
\newcommand{\daisyconf}[1]{
\defbibfilter{daisy-conf}{keyword=daisy and ( type=inproceedings or type=conference ) and not keyword=popular and not keyword=report}
\printbibliography[heading=subbibliography,%
filter=daisy-conf,title={#1}]
}
\newcommand{\daisyreport}[1]{
\defbibfilter{daisy-report}{keyword=daisy and ( type=report or keyword=report ) and not keyword=popular}
\printbibliography[heading=subbibliography,%
filter=daisy-report,title={#1}]
}
\newcommand{\daisyproject}[2]{
\defbibfilter{daisy-#1}{( keyword=daisy  or keyword=project ) and keyword=#1}
\printbibliography[heading=subbibliography,%
filter=daisy-#1,title={#2}]
}

\begin{document}

\title{Daisy literature}
\author{Per Abrahamsen}

\maketitle{}

This note is intended to provide an overview of papers that concerns
Daisy.  The papers are listed up to three times, first organized by type,
then organized by subject area, and finally organized by funding
project.  The papers are listed in reverse chronological order.  Only
papers that represent a signficant deployment or development of the
model are included. Still, most papers are likely missing.  Please
send missing references to daisy@daisy-model.dk.

\nocite{*}

\section*{Overview by type}

\daisytype{article}{Peer reviewed articles}
\daisytype{incollection}{Book chapters}
\daisyconf{Conference and workshop proceedings}
\daisytype{thesis}{Thesises and dissertations}
\daisyreport{Reports}
\daisykey{popular}{Popular articles}
\defbibfilter{daisy-other}{keyword=daisy and not keyword=popular and not type=article and not type=incollection and not type=inproceedings and not type=thesis and not type=report}
\printbibliography[heading=subbibliography,%
filter=daisy-other,title={Other sources}]

\section*{Overview by subject}

Daisy litterature divided by subject.  As some papers covers multiple
subjects, there will be some overlap.

\daisykey{description}{Model descriptions}
\daisykey{pesticides}{Pesticide fate}
\daisykey{climate}{Climate change}
\daisykey{rs}{Remote sensing}
\daisykey{svat}{Soil vegetation atmosphere transfer and evapotranspiration}
\daisykey{hydrology}{Soil hydrology}
\daisykey{usecase}{Use cases}
\daisykey{nleach}{Nitrogen leaching}
\daisykey{som}{Soil organic matter}
\daisykey{dss}{Decision  support}
\daisykey{farming}{Farming}
\daisykey{organic}{Organic farming}
\daisykey{crop}{Crop modelling}
\daisykey{intercrop}{Intercropping}
\daisykey{software}{Model software}
\daisykey{danish}{Danish language}

\section{Overview by project}

Daisy litterature divided by associated project.  As some papers
describe worked funded by multiple projects, there are some overlap.
Most papers have not yet been classified by project.

\daisyproject{mst-tiledrain}{MST-Markdræn}
\daisyproject{safir}{SAFIR}
\daisyproject{fertorganic}{FertOrgaNic}
\daisyproject{rsmodel}{RS-Model}
\daisyproject{sdss}{SDSS}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
