\documentclass[a4paper,11pt,twoside]{amsart}
\usepackage{a4}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{hyperref}
\usepackage{natbib}
\input{macro}

\begin{document}

\section*{Estimating the root density distribution from root dry matter and 
  depth}

\section{Uniform field crops}

In accordance with \cite{gp74}, the root density distribution $L_z$
for a crop can be described by
\begin{equation}
  L_z = L_0\, e^{-a z}
  \label{eq:g+p}
\end{equation}
where $L_0$ is the root density at the soil surface, $a$ is a
distribution parameter, and $z$ is the depth below soil surface.

We here assume that the density is uniformly distributed on the
horizontal plane, an assumption that fails with e.g.\ row crops.

The parameters $a$ and $L_0$ will both vary with time.  For a
production oriented simulation model like
Daisy~\citep{daisy-fertilizer,daisy-ems}, it can be more convenient to
specify the density in terms of accumulated root dry matter $M_r$ and
total root depth $d$, as described in \cite{daisy-def} or the
following.

We define the root depth at the lowest depth where the root density is
at above specified threshold $L_m$.  By inserting this in
(\ref{eq:g+p}), we get
\begin{equation}
  L_m = L_0\, e^{-a d}
  \label{eq:root-depth}
\end{equation}

We convert the root mass to root length $l_r$ by assuming the specific
root length $S_r$ is a known constant (rather than varying with depth)
\begin{equation}
  l_r = S_r \, M_r
  \label{eq:root-length}
\end{equation}

The total root length is also the integral of the root density over
the profile
\begin{equation}
  l_r = \int_0^{\infty} L_z \: dz = \int_0^{\infty} L_0\, e^{-a z} \, dz
  \label{eq:root-integral}
\end{equation}

An indefinite integral to (\ref{eq:g+p}) is 
\begin{equation}
  G_z = -\frac{L_0}{a}\, e^{-a z}
  \label{eq:indefinite}
\end{equation}

Using (\ref{eq:indefinite}) in (\ref{eq:root-integral}) gives us
\begin{equation}
    l_r = G_{\infty} - G_0 = \frac{L_0}{a}
  \label{eq:root-integrated}
\end{equation}

By inserting the expressin we get for $L_0$ from
(\ref{eq:root-integrated}) in (\ref{eq:root-depth}) we get
\begin{equation}
  L_m = L_r \, a \, e^{-a d}
  \label{eq:a-only}
\end{equation}

If we substitute $W = -a d$ (and thus $a = -W / d$) we get
\begin{equation}
   L_m = L_r \frac{-W}{d} e^W
  \label{eq:sub-W}
\end{equation}

By isolating the known values on the right side this gives us
\begin{equation}
   W e^W = - L_m \frac{d}{L_r}
  \label{eq:Lambert}
\end{equation}

The solution to this equation with regard to $W$ happens to be the
definition of the Lambert-W function \citep{euler83,lambert58}.  An
existing implementation such as \citep{lambertwcode} can then be used
to find a by substituting $a$ back
\begin{equation}
   a = -\frac{W (-\frac{L_m d}{L_r})}{d}
  \label{eq:a-solved}
\end{equation}

Having found $a$, $L_0$ can be determined by rewriting
(\ref{eq:root-depth}) into
\begin{equation}
  L_0 = \frac{L_m}{e^{-a d}}
  \label{eq:L0-found}
\end{equation}

\section{Limited growth}

The distribution (\ref{eq:g+p}) implies a gradual decrease of roots
going towards, but never reaching zero.  There are two problems with
this.  The first one is emprical, for some soils it doesn't match what
we observe, rather than a gradual decrease, there is sharp decrease at
a specific depth, as the roots are unable to penetrate further down.
The other one is practical, too large a root zone makes computation
impractical.

The way we model the first issue is to divide the root depth into a
crop specific and soil independent potential root depth $d$, and soil
specific and crop independent maximum root depth $d_s$.  The actual
root depth $d_a$ is then the shallowest of these two.
\begin{equation}
  d_a =
  \begin{cases}
    d & \text{if $d < d_s$}\\
    d_s & \text{if $d \geq d_s$}
  \end{cases}
  \label{eq:actual-depth}
\end{equation}

We now create a modified root density function $L_z^*$ by defining it to
zero below $d_a$, and a $L_z$ scaled to preserve mass balance above.
\begin{equation}
  L_z^* =
  \begin{cases}
    0 & \text{if $z > d_a$} \\
    k^* L_z & \text{if $z \leq d_a$}
  \end{cases}
  \label{eq:limited-depth}
\end{equation}
where
\begin{equation}
  k^* = \frac{l_r}{\int_0^{d_a} L_z \, dz}
\end{equation}
thus solving both problems.

\section{Row crops}

We can describe a row crop with a two dimensional model by assuming
that the plants are densily packed in the row.  Our second dimension
$x$ is horizontal, ortogonal to the row.  The root density at a
specific point can be denoted $L_{z,x}$, and we chooce origo so
$L_{0,0}$ is the the root density in the top of the row.  We also need
to know the distance between rows $R$, and assume an infinite number
of identical rows to avoid any border problems.

We would like to retain our original distribution when ignoring the x
dimension
\begin{equation}
  L_z = \frac{2 \int_0^{\infty} L_{z,x}\, dx}{R}
  \label{eq:x-integrated}
\end{equation}
We multiply by two as we assume the two sides of the rows are
identical.  By integrating to $\infty$ rather than just $R/2$ we do
include roots from outside the row.  However, because the system has
an infinite number of identical rows, the amount of roots from the
crop outside its own row is exactly the same as the amount of roots
from other rows inside the row we are examining.

If we further assumes that the root density decrease with the
horizontal distance can also be described with (\ref{eq:g+p}), we get
\begin{equation}
  L_{z,x} = L_{0,0}\, e^{-a_z z} e^{-a_x x}
  \label{eq:Lzx}
\end{equation}
Inserting (\ref{eq:Lzx}) and (\ref{eq:g+p}) in (\ref{eq:x-integrated}) we get
\begin{equation}
  \begin{array}{rcl}
    L_0\, e^{-a z} &=& \frac{2}{R} \int_{0}^{\infty} L_{0,0}\, e^{-a_z z} e^{-a_x x} dx\\
                 &=& \frac{2 L_{0,0}\, e^{-a_z z}}{R} \int_{0}^{\infty} e^{-a_x x} dx\\
                 &=& L_{0,0}\, e^{-a_z z} \frac{0 - 1}{-a_x}\\
                 &=& \frac{2 L_{0,0}}{R a_x} e^{-a_z z}\\
  \end{array}
  \label{eq:1d2d}
\end{equation}
So we get
\begin{eqnarray}
  a_z &=& a\\\label{eq:azisa}
  L_{0,0} &=& ½ a_x R L_0\\\label{eq:L00L0}
  L_0 &=& \frac{2 L_{0,0}}{a_x R}\label{eq:L0L00}
\end{eqnarray}
when switching between the one and two dimensional descriptions.

\section{Finding the parameters}

To find the parameters $a_z$, $a_x$ and $L_{0,0}$, we assume as before
that the root depth and root mass is known, and now additionally that
the root width $w$ is known.  As we only calculate on one side of the
row (the other being symetrical) we use the radius $r$ instead, where
$r = w/2$.

We used the minimum root density $L_m$ to define the width and depth
as before, but since it vary with the other dimension, we choose the
maximum point, which happens to be at the surface and at the root,
respectively
\begin{equation}
  L_m = L_{d,0} = L_{0,r}
  \label{eq:minroot}
\end{equation}

The total root length is the integral of the root density over the
half plane
\begin{equation}
  \begin{array}{rcl}
    l_r &=& \int_0^{\infty} \int_0^{\infty} L_{z,x} \: dz \, dx \\
        &=& \int_0^{\infty} \int_0^{\infty} e^{-a_z z} e^{-a_x x} \: dz \, dx\\
        &=& L_{0,0} \int_0^{\infty} e^{-a_x x} \int_0^{\infty} e^{-a_z z} \: dz \, dx\\
        &=& \frac{L_{0,0}}{a_z} \int_0^{\infty} e^{-a_x x} \:  dx\\
        &=& \frac{L_{0,0}}{a_z a_x}\\ 
  \end{array}
  \label{eq:root-integral2}
\end{equation}
Thus (\ref{eq:Lzx}) can be rewritten 
\begin{equation}
  L_{z,x} = l_r\, a_z\, a_x \, e^{-a_z z} e^{-a_x x}  
  \label{eq:azx}
\end{equation}

By using (\ref{eq:azx}) in (\ref{eq:minroot}) we get
\begin{eqnarray}
  L_m &=& l_r\, a_z\, a_x \, e^{-a_z d} \\\label{eq:Ld0}
  L_m &=& l_r\, a_z\, a_x \, e^{-a_x r}\label{eq:L0r}
\end{eqnarray}
Thus $e^{-a_z d} = e^{-a_x r}$ or
\begin{equation}
  a_x = \frac{d}{r} a_z
  \label{eq:aztoax}
\end{equation}

By inserting (\ref{eq:aztoax}) in (\ref{eq:Ld0}) we get
\begin{equation}
  L_m = l_r\, a_z\, \frac{d}{r} a_z \, e^{-a_z d}
  \label{eq:azeq1}
\end{equation}

If we substitute 
\begin{equation}
  Q = -a_z d
  \label{eq:Qaz}
\end{equation}
and isolate the known values on the right side, this gives us:
\begin{equation}
  Q^2\, e^Q = L_m \frac{d \, r}{l_r}
  \label{eq:logsquare}
\end{equation}

\section{Numeric solution}

Unlike (\ref{eq:Lambert}), nobody bothered to give the solution to
\ref{eq:logsquare} a name, so we have to solve it numerically.

We start by dividing the funtions into monotonic intervals by finding
the derivative
\begin{equation}
  \frac{d Q^2\, e^Q}{dQ} = 2 Q e^Q + Q^2 e^Q
  \label{eq:derived}
\end{equation}
The equation
\begin{equation}
  2 Q e^Q + Q^2 e^Q = 0
  \label{eq:derived-solutions}
\end{equation}
has two solutions, $Q=0$ and $Q=-2$, and the expression $Q^2\, e^Q$ is
increasing below $-2$, decreasing between $-2$ and $0$, and increasing
above $0$. Thus, $Q=0$ is a local (and in this case also global)
minimum, and $Q=-2$ is a local maximum.

We are not interested in positive values for $Q$, they correspond to
negative values for $a_z$, and the simplification in
(\ref{eq:root-integral2}) are only valid if $a_z > 0$.  

Since $\lim_{Q\to-\infty} Q^2\,e^Q=0$ we get a single negative
solution when $L_m \frac{d \, r}{l_r}$ is exactly at the top point
($2^2 e^{-2}$), two when it is smaller (it is never negative), and
none when it is larger.  The later situation corresponds to the case
where there are insufficent root $l_r$ to satisfy the minimal root
density $L_m$ within the given root zone $d\, r$. 

Both negative solutions are valid, but represent different
distributions.
\begin{itemize} 
\item The solution for $Q < -2$ represents a large $a_z$ (and thus
  also $a_x$) parameter. From (\ref{eq:root-integral2}) we see this
  also means $L_{0,0}$ is large.  Thus, the solution corresponds to a
  root zone with a high density near the center that decreases rapidly
  to $L_m$ near the edge of the root zone, and continues to decrease
  so only a small contribution to the total root length fro outside
  the root zone.
\item The solution for $Q > -2$ (and thus small values of $a_x$, $a_x$
  and $L_{0,0}$) corresponds to a low root density near the center
  that decreases slowly, and thus gives a larger contribution to the
  total root length from outside the root zone.  
\end{itemize}
As the total root length increases, pressing Q towards $0$ or
$-\infty$, the difference in the solutions grow.  When there is just
enough roots to satisfy the contraints at $Q = -2$, the two solutions
converges to one.  As we like our roots to stay mostly within the root
zone, we choose the solution for $Q < -2$.

We can find $Q$ numerically using Newton's method and an initial guess
of $-3$.  From that we can find $a_z$ from (\ref{eq:Qaz}), $a_x$ from
(\ref{eq:aztoax}), and $L_{0,0}$ from (\ref{eq:root-integral2}).

\section*{List of symbols}

\begin{tabular}{lll}
Symbol  & Unit    & Description\\\hline
$a$     & m$^{-1}$ & Root density distribution parameter\\
$a_z$   & m$^{-1}$ & Vertical root density distribution parameter\\
$a_x$   & m$^{-1}$ & Horizontal root density distribution parameter\\
$d$     & m       & Root depth\\
$d_a$   & m       & Soil limited root depth\\
$d_s$   & m       & Soil maximum root depth\\ 
$k^*$   &         & Soil root limit factor\\ 
$l_r$   & m/m$^2$ & Total root length\\ 
$L_0$   & m/m$^3$ & Average root density at soil surface\\
$L_{0,0}$& m/m$^3$ & Root density in row at soil surface\\
$L_m$   & m/m$^3$ & Minimal root density\\ 
$L_z$   & m/m$^3$ & Root density at soil depth $z$\\
$L_z^*$ & m/m$^3$ & Soil limited root density at soil depth $z$\\ 
$L_{z,x}$& m/m$^3$ & Root density at soil depth $z$ and distance $x$ from row\\
$M_r$   & kg/m$^2$& Total root dry matter\\ 
$Q$     &         & Substitution variable\\
$r$     & m       & Horizontal root radius\\
$R$     & m       & Distance between rows\\
$S_r$   & m/kg    & Specific root length\\ 
$w$     & m       & Root width\\ 
$W$     &         & Lambert W function\\ 
$x$     & m       & Horizontal distance from row \\
$z$     & m       & Soil depth \\
\end{tabular}

\bibliographystyle{elsart-harv}
\bibliography{daisy}

\end{document}
